---
title: "Preprocessing"
output: html_notebook
---

# Pre-processing
```{r}
library(tidyverse)
library(astsa)
library(scales)
```

```{r}
e_df = read_csv('../data/unprocessed/data.csv')
summary(e_df)
```


```{r}
e_df %>% head()
```
```{r}
e_df$date = as.Date(e_df$date)

e_df %>% group_by(date) %>% summarize(len = length(date)) %>% ggplot(aes(x=len)) + geom_histogram(bins=50,col='black', fill='white') + ggtitle('Histogram of amount of earthquakes per day') + xlab('number of earthquakes per day')
```

Fill in the missing dates with N/A values:
```{r}
#make a df of the complete dates from 2008-11-01 to 2022-09-26
dates = tibble(seq(as.Date("2008-11-01"), as.Date("2022-09-26"), by="days"))
colnames(dates) = 'date'

#make sure each date is available, and fill with NA
newdf = merge(e_df, dates, by='date',all.y = TRUE, sort=TRUE)

missing_dates = c(newdf %>% filter(is.na(magnitude)) %>% select(date))

missing_percent = percent(length(missing_dates$date)/length(dates$date))

cat('Percentage of dates that have N/A values:', missing_percent, 'of', length(dates$date),'dates.')

#take average of adjacent earthquakes in time
```

Because of this excess in missing dates, let's take the monthly average. I have tried running it daily with this much missing data and there's lots of problems.

```{r}
more1count = e_df %>% group_by(date) %>% summarize(len = length(date)) %>% filter(len>1) %>% nrow()
perc1count = percent(more1count/length(dates$date))

cat('Percentage of dates that have more than 1 earthquakes:', perc1count, 'of', length(dates$date),'dates.')
```

# Feature Engineering
Creating year and quarters and month_year variables:
```{r}
newdf = newdf %>% 
  
  #getting the year from date
  mutate(year = as.integer(format(date,'%Y'))) %>%
  
  #getting the month from date
  mutate(month = as.integer(format(date,'%m'))) %>%
  
  #creating magnitude categories
  mutate(quarter = factor(case_when(month < 4 ~ 1,
                             month < 7 ~ 2,
                             month < 10 ~ 3,
                             month < 13 ~ 4), levels=c(1,2,3,4))) %>%
  
  
  #getting month_year
  mutate(y_m = paste0(year, '-',month)) %>% 
  mutate(mag_cat = factor(case_when(magnitude < 2.5 ~ 'very weak',
               magnitude <= 5.4 ~ 'weak',
               magnitude <= 6 ~'medium',
               magnitude <= 6.9 ~ 'strong',
               magnitude <= 7.9 ~ 'very strong',
               magnitude > 7.9 ~ 'extreme')))


newdf %>% head()
```

Distribution of count values of month_year
```{r}
newdf %>% group_by(y_m) %>% summarize(count = n()) %>% ggplot(aes(x=y_m,y=count)) + geom_bar(stat='identity',aes(fill=count)) + ggtitle('Distribution of count in each month_year')
```
Finding out what's the largest count
```{r}
newdf %>% group_by(y_m) %>% summarize(count = n()) %>% filter(count == max(count))
```

Distribution of magnitude in each month_year
```{r}
newdf %>% group_by(y_m) %>% summarize(mag_max = max(magnitude)) %>% ggplot(aes(x=y_m,y=mag_max)) + geom_bar(stat='identity',aes(fill=mag_max)) + ggtitle('Distribution of magnitude in each month_year')
```

The month_year 8 2018 has 2297 counts
The lombok earthquake

```{r}
cat('The number of unique month_year is:', newdf %>% select(y_m) %>% unique() %>% nrow())
```

The dataframe we're going to use
```{r}
m_df = newdf %>% group_by(y_m) %>% filter(magnitude == max(magnitude, na.rm=TRUE))
m_df
```
I noticed that there are missing y_m in the data.

List of total y_m and merge with m_df
```{r}
list_my = newdf %>% select(y_m) %>% unique() %>% select(y_m)

#make sure each date is available, and fill with NA
m_df = merge(m_df, list_my, by='y_m',all.y = TRUE, sort=TRUE)

m_df %>% filter(is.na(magnitude)) %>% pull(y_m)
```
Wow! These months have no recorded earthquakes. Maybe because of recording error or holidays.

For SARIMA, we need to impute these!
Let's get the list of yearly average and impute these with a bit of noise.
```{r}
#get yearly averages
year_avg = m_df  %>%  
  group_by(year) %>%
  summarize(y_mean = mean(magnitude,na.rm=TRUE))

year_avg %>% ggplot(aes(x=year,y=y_mean, fill=y_mean)) + geom_bar(stat='identity') + ggtitle('Yearly averages of earthquakes')
```
```{r}
ggplot(na.omit(m_df),aes(x=longitude, y=latitude)) + 
  geom_point(shape=24,aes(fill=mag_cat,alpha=mag_cat)) + 
  scale_alpha_manual(values=c(1,0.5,0.4,0.3,0.2,0.1)) + ggtitle('Spatial plot of earthquakes')
```
use year_avg to impute the data
```{r}
#for reprodudibility
set.seed(123)

#get year using regex
m_df = m_df %>% mutate(year = as.integer(sub('-(.+)', '',y_m))) %>% 
  mutate(month = as.integer(sub('(.+)-','',y_m)))

#impute the data
m_df_imputed = m_df %>% mutate(magnitude = ifelse(is.na(magnitude), year_avg %>% filter(year==year) %>% pull(y_mean) + rnorm(1,mean=0,sd=1),magnitude)) %>%   
  mutate(mag_cat = factor(case_when(magnitude < 2.5 ~ 'very weak',
               magnitude <= 5.4 ~ 'weak',
               magnitude <= 6 ~'medium',
               magnitude <= 6.9 ~ 'strong',
               magnitude <= 7.9 ~ 'very strong',
               magnitude > 7.9 ~ 'extreme')))
```


```{r}
write_csv(m_df, file='../data/processed/month_df.csv')
write_csv(m_df_imputed, file='../data/processed/month_df_imputed.csv')
```

